---
title: "Assignment5_BurialMoundHealth"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 2: Where in the landscape are the mounds located?

The location of settlements is usually easy to predict as humans need safety and easy access to water and food resources. These range from nearby arable soil to pasturage to trading centers. Choices regarding the mortuary realm are much harder to establish as they are guided by the social norms of the living rather than the natural environment. Some environmental conditions, however, play a role, such as intervisibility, which can be an important factor for societies that use monuments to the dead for territorial signalling. Before such specific analysis, it is, however, a good idea to get a general sense of where in the landscape are mounds located.

In order to produce a formal assessment of mound location, you can start by using most common aspects of topography, such as elevation, slope, aspect, and water proximity. Choose one or more of these variables. Calculate the distribution of classes of each environmental variable through the entire region (defining, en route, what you consider to be a “region”?) by extracting values from the digital elevation model and binning them in several classes. Then, calculate site frequencies within classes of the environmental variable by sampling mound locations in the raster and evaluate whether the observed pattern could be a product of the distribution of environmental categories themselves.

An example workflow with elevations could look like this:
1. Extract elevation values from the entire landscape and bin them in elevation categories (e.g. 400-500m, 500-600m, 600-700m, etc.). Consider: what defines a landscape for the purpose of this study? You may wish to crop the Aster to a smaller but representative study area but make sure you justify your selection
2. Extract elevation values from observed mound locations and review their distribution in the same bins
3. Calculate the expected number of mounds per bin if mounds were located through the landscape randomly
4. Compare the expected number with the observed one

__Packages__
```{r}
library(tidyverse)
library(raster)
library(sf)
library(lattice)
library(latticeExtra)
library(RColorBrewer)
library(rasterVis)
library(rgdal)
library(rgl)
library(tmap)
```

__Data__
For this task I will need the digital elevation for the Kazanlak Valley with 30m resolution, which is the "Aster.tif" file. I will also be needing the shapefile containing the GPS points for each mound in the Kazanlak valley which is stored in the "Kaz_mounds.shp" file. 

```{r}
mounds <- st_read("data/KAZ_mounds.shp")
aster <- raster("data/Aster.tif")
```

__Examining Elevation__
To get a general sense of where in the landscape the mounds are located, I will first examine the elevation for the region of interest, which is the region in which the mounds are located. 

Before beginning to extract values from the raster, I need to make sure that the data share the same coordinate reference system (CRS), and if not I need to project to a common CRS.

```{r}
st_crs(mounds) # WGS84
crs(aster) # WGS84
```

Since they share a common CRS I will not need to reproject. 

Checking the satellite imagery by plotting:
```{r}
plot(aster)
```

This does not look right. From the plot I can see that there is a need to reclassify values that are between -10.000 and -1 to NAs. Missing values are commonly coded as -9999 and this is why they are scewing the rendering of the raster data, and we need to fix this with reclassification. 

```{r}
rclmat <- cbind(-10000,-1, NA) # specifying what we want to reclassify
aster_reclass <- reclassify(aster, rcl = rclmat) # reclassifying the aster data
plot(aster_reclass) # plotting to make sure we fixed the issue
```

Now the satellite imagery looks much better and we can continue. 
Now we can make a 3D plot using the rgl package to get a better sense of what the landscape looks like. 

```{r}
rglwidget()
plot3D(aster_reclass)
```

Before extracting values from the landscape, we need to crop it to a smaller, but more representative region. It makes sense to crop the landscape to only contain the area in which the burial mounds are located, so that it what we will do. We can use the crop() function to crop the landscape to only include the region in which the mounds are located.

```{r}
cropped_aster <- crop(aster_reclass, mounds)
```

Let's plot the cropped region.

```{r}
plot(cropped_aster) # simple plot
plot3D(cropped_aster) # 3D plot

# Plotting with tmap
cropped <- tm_shape(cropped_aster) +
  tm_raster() +
  tm_shape(mounds) +
  tm_markers(size = 0.3) +
  tm_layout(main.title = "Cropped Landscape and Markers for Mounds",
            main.title.size = 1)

cropped
```

Let's look at the cropped vs. uncropped landscape next to each other

```{r}
# Plotting the entire, uncropped landscape with tmap
uncropped <- tm_shape(aster_reclass) +
  tm_raster() +
  tm_shape(mounds) +
  tm_markers(size = 0.3) +
  tm_layout(main.title = "Uncropped Landscape and Markers for Mounds",
  main.title.size = 0.9)

# Plotting the cropped and uncropped next to each other
tmap_arrange(uncropped, cropped, asp = NA)
```

Now we can extract the elevation values from the landscape data.
```{r}
# Defining bins
breaks <- c(200, 300, 400, 500, 600, 700, 800, 900, 1000, 1100, 1200, 1300, 1400)

# Extracting elevation values from the entire, uncropped landscape
elevation_bins_uncropped <- cut(values(aster_reclass), breaks = breaks, dig.lab = 5) # dig.lab = 5 means 5 digits
elevation_df <- data.frame(table(elevation_bins_uncropped)) # converting the elevation values into a data frame
elevation_df 
# Now we have a dataframe containing the elevation values grouped in bins and how many mounds fall into each bin of elevation values which provides us with a nice overview of the distribution of elevation values for the entire landscape.

# Let's examine the distribution of elevation values for the entire cropped area
aster_cropped_elev <- values(cropped_aster) # extracting the elevation values for the cropped landscape
hist(aster_cropped_elev, breaks = breaks) # plotting 
```

For the cropped landscape area we can see that most mounds are within the 300-400 elevation values bin. 
Now we can move on to examining the distribution of elevation values for the mounds in particular. 

Fitst we extract the elevation values for the cropped region that contains the burial mounds and with a histogram we examine the distribution of elevation values of the mounds. 
```{r}
# Extracting the elevation values for burial mounds in the cropped area
mounds$elevation <- raster::extract(cropped_aster, mounds)

# Examine the distribution of elevation values in the cropped area in which the burial mounds are located
hist(mounds$elevation, breaks = breaks)
```

In contrast to the entire cropped area in which most mounds were in the 300-400 bin of elevation values, the elevation values only for the burial mounds are wihtin the 400-500 elevation value bin. 
Let's look at the exact frequency of mounds per bin.

```{r}
mounds_freq <- cut(mounds$elevation, breaks = breaks, dig.lab = 5)
mounds_freq <- data.frame(table(mounds_freq))
mounds_freq
```

Now we have the observed freuqnecies per bin, and we know that most burials mounds (616) are located in areas with an elevation between 400 and 500. Now we can calcualte the number of mounds per bin that would be expected if mounds were RANDOMLY located through the landscape randomly. We do so by taking a random sample of mounds from the cropped region and extract their elevation values and examine the distribution of these elevation values to compare with the expected elevation values we found above.  

```{r}
set.seed(1) # set seed to get the same values each time we take a sample
random <- sampleRandom(cropped_aster, 773) # taking a random sample of mounds from the cropped area
random$elevation = raster::extract(cropped_aster, random) # extracting the elevation values from the random sample
hist(random$elevation, breaks = breaks) # examining the elevation value distribution for the random sample
random_freq <- cut(random$elevation, breaks = breaks, dig.lab = 5) # calcuating the expected frequency per bin
random_freq <- data.frame(table(random_freq)) # converting to a dataframe
random_freq
```
From the above we can conclude that if the burials mounds were randomly located throughout the landscape, most of them (416) would be located between 500-600 in elevation. Since this distribution does not correspond to what we actually observe, we can infer that the burial mounds are most likely NOT randomly located. 

Now we have looked at elevation to assess the location of burial mounds, and now we turn to look at the slope for the burial mounds and see if this can tell us more about why the burials mounds are located where they are. 

__Examining Slope__
Below we are going to look at the slopes of the burial mounds to see if the slopes of the burials mounds seem random or if there is a particular pattern present. Hence, we basically repeat what we did above with the elevation data. 

First we need to extract the slope values for the burial mounds
```{r}
slopes_cropped <- terrain(cropped_aster, opt = "slope", unit = "degrees") # extracting slopes
plot(slopes) # simple plot
plot3D(slopes) # 3D plot
```

Now we can define the bins for the slope values and examine the distribution of mounds according to slope.
```{r}
# Define bins
breaks <- c(0, 10, 20, 30, 40, 50, 60)

# Extract slope values
slopes_bins <- cut(values(slopes_cropped), breaks = breaks, dig.lab = 5) # dig.lab = 5 means 5 digits
slopes_df <- data.frame(table(slopes_bins)) # converting the slope value table into a data frame
slopes_df

# Histogram
slopes_values <- values(slopes_cropped)
hist(slopes_values, breaks = breaks)
```

We can see that most of the burial mounds wihtin the cropped area are located in an area with a slope between 0-10 degress, hence, not very steep. 

Now let's examine the slope for the particular mounds we are interested in:
```{r}
mounds$slope <- raster::extract(slopes_cropped, mounds)

# Examine the distribution of elevation values in the cropped area in which the burial mounds are located
hist(mounds$slope, breaks = breaks)

# Extract the actual frequency
mounds_freq_slope <- cut(mounds$slope, breaks = breaks, dig.lab = 5)
mounds_freq_slope <- data.frame(table(mounds_freq_slope))
mounds_freq_slope
```

Most burials mounds (728) are located in areas with a slope between 0-10 degrees.

Now we can calculate the expected number of mounds per bin if the mounds were located randomly through the landscape, and see how this compares to what we actually observe.

```{r}
set.seed(1) # set seed to get the same values each time we take a sample
random <- sampleRandom(cropped_aster, 773) # taking a random sample of mounds from the cropped area
random$slope <- raster::extract(slopes_cropped, random) # extracting the elevation values from the random sample
hist(random$slope, breaks = breaks) # examining the elevation value distribution for the random sample
random_freq <- cut(random$slope, breaks = breaks, dig.lab = 5) # calcuating the expected frequency per bin
random_freq <- data.frame(table(random_freq)) # converting to a dataframe
random_freq
```

With a random sample of burial mounds, we can see that most would be located in areas with slopes between 0-10 or between 20-30 degrees. This suggest that it might not be entirely random that the burial mounds in the Kazanlak valley are located in areas with a slope between 0-10, since a random sample also shows that many mounds (16) would be expected to have a slope between 20-30. 

__Conclusion__
Based on an assessment of common aspects of topographic, elevation and slope, we can conclude that the burial mounds in the Kazanlak Valley are most likely not located randomly, rather, there seems to be particular patterns when it comes to elevation and slope for these particular burials mounds. 
