---
title: "Country and the city: Danish voters and party preferences"
date: "19/03/2021 updated `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
---


```{r setup, include=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r libraries, include=FALSE}
# Library 
library(raster)
library(rgeos)
library(sf)
library(tidyverse)
library(htmltools)
library(googlesheets4)
library(mapview)
```

# Task 1: Get spatial data for municipalities in Denmark 
You can download administrative data for Denmark from the GADM dataset, the Global administrative boundaries, hosted by UCDavis. You do this by using the getData() function in the `raster` package. For GADM data, you need to specify what level of admin boundaries you wish to download (0=country, 1=first level subdivision aka regions, 2=second level aka municipalities, etc.). Read [this blog](https://www.gis-blog.com/r-raster-data-acquisition/) on the power of `raster` package when it comes to available datasets. 

## Instructions:

* Use getData() function and `level = 2` to download the boundaries of Danish kommunes. 
* Note the class of the object and  variable called `NAME_2`, it's the name of the municipality
* Convert the Spatial dataframe to sf object and project UTM32
* Use `mapview` or `tmap` library to see a map with all the municipalities.

```{r load-spdata}
# Load the spatial data, project to UTM
# mun_sp<- getData('GADM', country = 'DK', level = 2) # this is for downloading the dataset
mun_sp <- readRDS("data/gadm36_DNK_2_sp.rds") # spatial dataframe
mun_sf <- st_as_sf(mun_sp) # converting to an sf-object 
mun <- st_transform(mun_sf, crs = 32632) # transforming to the Danish UTM projection
mapview(mun)

# Straighten the names. Some of the municipalities are not spelled in the same way. We need to fix that. 
sort(mun$NAME_2)
which(grepl("Å",mun$NAME_2))
which(grepl("Vest",mun$NAME_2))

mun$NAME_2[31] <- "Aarhus"
mun$NAME_2[21] <- "Høje-Taastrup"
mun$NAME_2[60] <- "Vesthimmerlands"
```

# Task 2: Wrangle the attributes and join to the spatial data
In order to show something we need to connect the spatial polygons with some attributes. Let's pick the civil status table from Denmark Statistik and calculate the total numbers of men and women in each municipality and calculate the percentages for single men and women so that the singles know where to go to find a significant other :). 

Here we get to practice basic tidyverse functions and the use of `tmap` package.

```{r load-elec}
# Load the attributes
#elec <- read_sheet("https://docs.google.com/spreadsheets/d/1ty3UrUiCK2iqWVk2T2GORaCl0QZ6feUTMmTakXPVIIg/edit#gid=0", col_types = "ccnnn")
#write_csv(elec, "data/elections.csv")

# Load file 
elec <- read_csv("data/elections.csv")

# Check names
sort(unique(elec$Region))

# Check out the data
# Summarize the total number of votes for each party across year
elec %>% 
  group_by(Party) %>% 
  summarize(sum2011 = sum(Y2011),
            sum2015 = sum(Y2015),
            sum2019 = sum(Y2019))  %>%
  janitor::adorn_totals(where = "row")

# Create total number of votes per municipality (region)
electorate <- elec %>% 
  group_by(Region) %>% 
  summarize(sum2011 = sum(Y2011),
            sum2015 = sum(Y2015),
            sum2019 = sum(Y2019)) 

# Merge the summary (electorate) with the granual election dataset and spatial polygons
elections <- mun %>% 
  dplyr::select(NAME_2) %>% # name_2 = region (municipalities)
  merge(elec, by.x = "NAME_2",by.y ="Region") %>%
  merge(electorate, by.x = "NAME_2",by.y ="Region") %>% 
  group_by(NAME_2, Party) %>% 
  mutate(pct_vote2011 = Y2011/sum2011*100, # calculate percetage of votes 
         pct_vote2015 = Y2015/sum2015*100,
         pct_vote2019 = Y2019/sum2019*100)
elections

# Map some aspect of the result to see no counties are missing
elections %>% 
  group_by(NAME_2) %>% # group by regions
  filter(grepl("^A", Party)) %>%  # filtering for Socialdemokratiet, which starts with A. 
  select(pct_vote2015) %>% 
  mapview()
# Here we see the percentage of votes in 2015 for Socialdemokratiet (dark = low, bright = high)
# We can change the year, to see how the votes change across time

# Save for later?
#write_rds(elections, "data/elections_sp.rds") 
```

# Task 3: Look at some of the data
```{r map-data}
# Let's map the two most popular parties, SD and Danske Folkeparti through time
library(tmap)

# Looking at percentage of votes for Socialdemokratiet and Dansk Folkeparti in 2011
elections %>% 
  filter(grepl("^A|^O",Party)) %>% 
  tm_shape() + 
  tm_facets("Party", ncol = 2) +
  tm_polygons("pct_vote2011",
              title= "Percentage of Votes in 2011")

# 2015
elections %>% 
  filter(grepl("^A|^O",Party)) %>% 
  tm_shape() + 
  tm_facets("Party") +
  tm_polygons("pct_vote2015",
              title= "Percentage of Votes in 2015")
# 2019
elections %>% 
  filter(grepl("^A|^O",Party)) %>% 
  tm_shape() + 
  tm_facets("Party") +
  tm_polygons("pct_vote2019",
              title= "Percentage of Votes in 2019")

# We can see the reaction in terms of votes to the Syrian refugees. The right wing (Dansk Folkeparti) gets more votes. 
```

# Task 4: Cartogram
As you can see from the maps, the area of municipalities varies considerably. When mapping them, the large areas carry more visual "weight" than small areas, although just as many people or more people live in the small areas. Voters in low-density rural regions can thus visually outweigh the urban hi-density populations.

One technique for correcting for this is the cartogram. This is a controlled distortion of the regions, expanding some and contracting others, so that the area of each region is proportional to a desired quantity, such as the population. The cartogram also tries to maintain the correct geography as much as possible, by keeping regions in roughly the same place relative to each other.

The `cartogram` package contains functions for creating cartograms. You give it a spatial data frame and the name of a column, and you get back a similar data frame but with regions distorted so that the region area is proportional to the column value of the regions.

You'll also use the sf package for computing the areas of newly generated regions with the `st_area()` function.

## Instructions

The `elections` sf object should be already loaded in your environment.

* Load the `cartogram` and `rgeos` packages.
* Plot total electorate against area for each region. Deviation from a straight line shows the degree of misrepresentation.
* Create a cartogram scaling to the "`pct_vote2011`" column.
* Check that the DF voter population is proportional to the area.
* Plot the "pct_vote2011" percentage on the cartogram. Notice how some areas have relatively shrunk or grown.


### Solution
```{r cartogram-DF, echo=FALSE}
# Use the cartogram package
library(cartogram)
library(sf)

# Let's look at Danske Folkeparti in 2015
# We subset for Dansk Folkeparti
DK_DF <- elections %>% 
  filter(grepl("^O",Party)) 

# Make a scatterplot of municipality electorate versus area and Danske Folkeparti voters and municipality area. 
names(DK_DF)
plot(DK_DF$sum2015, st_area(DK_DF, byid = TRUE)) # eligible voters agains size of region. There is a lot of variation. Cartogram can reduce this - make less variation. 
plot(DK_DF$pct_vote2015, st_area(DK_DF, byid = TRUE))

# Cartograms can help us reduce the variation we see in the plots above. Cartograms are not geographically correct - they are simulated, disorted spatial representation of municipalities in terms of number of votes (hence, the map of Denmark will look weird)

# Make a cartogram, scaling the area to the electorate
# Here we use the cartogram to scale the size of the region to the number of eligible voters
electorate2015 <- cartogram_cont(DK_DF, "sum2015")

# Now check the linearity of the electorate per municipality plot
plot(electorate2015$sum2015, st_area(electorate2015, byid = TRUE))
# The y-axis is the area of the region (municipality)

# Make a cartogram, scaling the area to the percentage of DF voters
DF2015 <- cartogram_cont(DK_DF, "pct_vote2015")
# Now the sizes of the regions correspond to the number of voters in percentage. 
# The cartogram_cont() makes simulations which is why it does not completely get rid of variation

# Check the linearity of the DF voter percentage per municipality plot
plot(DF2015$pct_vote2015, st_area(DF2015, byid = TRUE))
# Much more linear than the first plots we did (before scaling the area to the number of votes)
# Hence, adjusting the size of the region to correspond to the number of votes makes it more linear. 

# Make a fairer map of the DF voter percentage in 2015
plot(electorate2015$geometry, 
     col = "beige",
     main = "Electorate in DK municipalities 2015")
plot(DF2015$geometry,
     col="pink",
     main = "% of Danske Folkeparti voters across DK in 2015")

# We can use these map to compare regions of interest and see how the votes change in 2015 from total number of votes to percentage
```
Copacetic cartogram! Now try to rerun the cartogram for the Social Democrats in 2015
```{r carto-SD}
# Let's look at Social Democrats in 2015
DKSD <- elections %>% 
  filter(grepl("^A",Party)) 

# Make a cartogram, scaling the area to the total number of votes cast in 2015
electorate2015 <- cartogram_cont(DKSD, "sum2015")

# Now check the linearity of the total voters per municipality plot
plot(electorate2015$sum2015, st_area(electorate2015, byid = TRUE))

# Make a cartogram, scaling the area to the percentage of SD voters
DF2015 <- cartogram_cont(DKSD, "pct_vote2015")

# Check the linearity of the SD voters percentage per municipality plot
plot(DF2015$pct_vote2015, st_area(DF2015, byid = TRUE))


# Make a adjusted map of the 2015 SD voters
plot(electorate2015$geometry, 
     col = "beige",
     main = "Electorate in DK municipalities 2015")
plot(DF2015$geometry,
     col="pink",
     main = "% of Social Democrat votes across DK in 2015")
```


# Task 5: Spatial autocorrelation test
If we look at the facetted tmaps the election results in 2015 seem to have spatial correlation - specifically the percentage of voters favoring Danske Folkeparti increases as you move towards the German border. This trend is not as visible in the cartogram, where the growth is more apparent in Sjæland, and other islands, like Samsø. 
How much correlation is there, really?
By correlation, we mean : pick any two kommunes that are neighbors - with a shared border - and the chances are they'll be more similar than any two random boroughs. 
This can be a problem when using statistical models that assume, conditional on the model, that the data points are independent.

The `spdep` package has functions for measures of spatial correlation, also known as spatial dependency. Computing these measures first requires you to work out which regions are neighbors via the `poly2nb()` function, short for "polygons to neighbors". The result is an object of class `nb`. Then you can compute the test statistic and run a significance test on the null hypothesis of no spatial correlation. The significance test can either be done by Monte-Carlo or theoretical models.

In this example you'll use the Moran "I" statistic to test the spatial correlation of the Danske Folkeparti voters in 2015.

## Instructions I - defining neighbors

* Load the `elections` spatial dataset with attributes
* Consider simplifying the boundaries if the data is too heavy for your computer and takes long to visualise
* Load the spdep library and create nb object of neighbors using queen adjacency
* Pass `elections` to `poly2nb()` to find the neighbors of each borough polygon. Assign to `nb`.
* Get the center points of each borough by passing `elections` to `st_centroid` and then to `st_coordinates()`. Assign to `mun_centers`.
* Update the basic map of the DK municipalities by adding the connections.
  - In the second plot call pass `nb` and `mun_centers`.
  - Also pass `add = TRUE` to add to the existing plot rather than starting a new one.

```{r}
# Reload the data if needed
# elections <- readRDS("data/elections_sp.rds")

# Checking that it is all looking good
elections
plot(elections$geometry)
mun
plot(mun$geometry)


# Here we simplify the municipalities with a tolerance of 250 meters
# st_cast() is used to transform from geometry to polygons, because the st_simplify has changed it to geometry and we want it to be in polygons
mun_sm <- st_cast(st_simplify(mun, dTolerance = 250),
                     to = "MULTIPOLYGON")
# Plot
plot(mun_sm$geometry)

# Use the spdep package
library(spdep)

# We need to define the neighborsing municipalities 
# We choose to define neighborhood according to adjacency (touching borders)
# Make neighbor list following queen adjacency
nb <- poly2nb(mun_sm$geometry) 
nb # this is a list of all neighbors. We can see that there are 8 regions with no neighbors (these are the islands). This might have some implications for the weighted matrix we make later and consequently Moran's I.

# Get center points of each municipality in order to visualize the neighborhoods
mun_centers <- st_coordinates(st_centroid(mun_sm$geometry))

# Show the connections
# First we plot the borders (geometry) and then we plot the centroids
plot(mun_sm$geometry); plot(nb, mun_centers, col = "red",add = TRUE)
# Now we can see the neighborhoods. Only polygons that touch have neigbors, because we are using the adjacency definiton of neighborhood.
```


## Instructions II - Moran's I

Now that your neighbors are determined and centroids are computed, let's continuing with the Moran's I statistic

* Create a subset with municipalities for `O.Danske Folkeparti` 
* Feed the `pct_2011` vector into `moran.test()`.
  - `moran.test()` needs a weighted version of the `nb` object which you get by calling `nb2listw()`. 
  - After you specify your neighbor `nb`object (`mun_nb`) you should  define the weights `style = "W"`. Here, `style = "W"` indicates that the weights for each spatial unit are standardized to sum to 1 (this is known as row standardization). For example, municipality 1 has 3 neighbors, and each of those neighbors will have weights of 1/3.   This allows for comparability between areas with different numbers of neighbors.
  - You will need another argument in both spatial weights and at the level of the test. `zero.policy= TRUE` deals with situations when an area has no neighbors based on your definition of neighbor (many islands in Denmark). When this happens and you don’t include `zero.policy= TRUE`, you’ll get the following error
  - Run the test against the theoretical distribution of Moran's I statistic. Find the p-value. Can you reject the null hypothesis of no spatial correlation?
* Inspect a map of `pct_2011`.
* Run another Moran I statistic test, this time on the percent of single women.
  - Use 999 Monte-Carlo iterations via `moran.mc()`.
  - The first two arguments are the same as for `moran.test()`.
  - You also need to pass the argument `nsim = 999`.
  - Note the p-value. Can you reject the null hypothesis this time?

```{r Moran-DF}
# With Moran's I we are checking whether neigbors are more correlated in terms of voting preferences compared to a random distribution 

# Load data
elections <- read_rds("data/elections_sp.rds")

# Let's look at Danske Folkeparti in 2015
DKF <- elections %>% 
  filter(grepl("^O",Party))

# Run a Moran I test on Dansk Foleparti percentage in 2015
moran.test(DKF$pct_vote2015,
           nb2listw(nb, style = "W",zero.policy=TRUE), # nb2listw creates a weighted list of neighbors. Style W = each neighbor is weighted the same.
           zero.policy=TRUE) #zero.policy is something we use if there are regions with no neighors. Zero.policy replaces these regions with a 0. 

# NB! The two objects that you are testing (pct_vote2015 and nb in this case) need to have the same number of rows otherwise the moran.test function won't work

# Remember that you cannot trust the p-value from the moran.test function - wait to make any conclusions until you have made the Monte Carlo simulation.
# Positive Moran's I: clustering
# Negative Moran's S: dispersal

# Run a Moran I test on Dansk Foleparti percentage in 2011
moran.test(DKF$pct_vote2011, 
           nb2listw(nb, style = "W",zero.policy=TRUE),
           zero.policy=TRUE)

# Do a Monte Carlo simulation to get a better p-value
moran.mc(DKF$pct_vote2015,nb2listw(nb, zero.policy=TRUE),zero.policy=TRUE, nsim = 999) # here we are randomly distributing the values for 999 simulations 
# Now we can actually look at the p-value. It is above 0.05 which means that there is a chance that the distribution is random. The correlation is positive (clustering)

# The p-value is the threshold for when we reject the null hypothesis. The p-value indicates the chance for the results deriving from a random distribution. Hence, the lower the p-value is, the lower the chance is that the results we are getting are random - hence, there is a high chance that they are not random (which means that there is some kind of spatial correlation)
```

Marvellous Moran Testing! You should have found that the p-value was around 0.079 in 2015 and 0.15 in 2011 the first case, thus you did not find any significant spatial correlation. In Monte Carlo simulation, the p-value was around 0.053, so you did find some not very significant spatial correlation (strongly positive).

### Repeat the same test for Social Democrats
```{r Moran-SD}
# Let's look at Social Democrats
DKSD <- elections %>% 
  filter(grepl("^A",Party))

# Run a Moran I test on total population of women
moran.test(DKSD$pct_vote2015, 
           nb2listw(nb, style = "W",zero.policy=TRUE),
           zero.policy=TRUE)

# Run a Moran I test on the proportion of single women
moran.test(DKSD$pct_vote2011, 
           nb2listw(nb, style = "W",zero.policy=TRUE),
           zero.policy=TRUE)

# Do a Monte Carlo simulation to get a better p-value
moran.mc(DKSD$pct_vote2011,nb2listw(nb, zero.policy=TRUE),zero.policy=TRUE, nsim = 999)

# Do a Monte Carlo simulation to get a better p-value
moran.mc(DKSD$pct_vote2015,nb2listw(nb, zero.policy=TRUE),zero.policy=TRUE, nsim = 999)
```

Phenomenal political testing. Social Democrats show even less correlation. P-value in Moran I test is was around 0.13 in 2011 results and 0.24 in 2015 results, thus no significant spatial correlation. In Monte Carlo simulation, the p-value was around 0.24, suggesting there is insignificant (positive) spatial correlation.

Well-done! Not so much correlation as it might seem at the first sight.